Dark Control:\\ A Unified Account of Default Mode Function\\ by Control Theory and Reinforcement Learning

### 2017-09-08 paper/paper.tex
Add Venn Diagram figure + Typos corrections

\subsubsection{Optimal control via stochastic gradient descent in the DMN}
Efficient learning of the entire set of model parameters can effectively be achieved
via stochastic \textit{gradient descent}, a universal algorithm for finding
local minima based on the first derivative of the optimization objective.
Stochastic here means that the true gradient is estimated from batches of training examples samples,
which, in our case, corresponds to blocks of experience from the replay memory:

@@

\section{Relation to existing accounts}
\subsection{Predictive coding hypothesis}
Predictive coding mechanisms
\citep{clark2013whatever, friston2008hierarchical}
are a frequently evoked idea in the context of default mode function
\citep{bar2007}.
Cortical responses are explained as
emerging from continuous functional interaction between
higher and lower levels of the neural processing hierarchy.
Feed-forward sensory processing is constantly calibrated by
top-down modulation from more multi-sensory and associative brain regions
further away from primary sensory cortical regions.
The dynamic interplay between cortical processing levels
may enable learning about world aspects by reconciling
gaps between fresh sensory input and expectations computed
based on stored prior information.
At each stage of neural processing,
an internally generated prediction of aspects of environmental sensations is
directly compared against the actual environmental input.
A prediction error at one of the processing levels
incurs plasticity changes of neuronal projections
(i.e., adapting model parameters)
to allow for gradually improved future prediction of the environment.
In this way,
the predictive coding hypothesis offers explanations for
the constructive, non-deterministic nature of sensory perception
\citep{friston2010free, buzsaki2006rhythms} and
the intimate relation of motor action to sensory expectations
\citep{wolpert1995internal, kording2004bayesian}.
Contextual integration of sensorimotor perception-action cycles
may be maintained by top-down modulation
using a-priori information about the environment.

In short, predictive coding processes
conceptualize updates of the internal representation of the environment
to best accommodate and prepare the organism
for processing the constant influx of sensory stimulation and
performing action on the environment.
There are hence a number of common properties
between the predictive coding account
and the proposed formal account of DMN function based on MDPs.
Importantly,
a generative model of how perceived sensory cues arise in the
world would be incorporated into
the current neuronal wiring.
Further,
both functional accounts are plausibilized by
neuroscientific evidence that suggest
the human brain to be a ``statistical organ'' \citep{friston2014phantastic}
with the biological purpose to
generalize from the past to new experiences.
Neuroanatomically, axonal
back projections indeed outnumber by far the axonal input projections
existing in the monkey and probably also human brain
\citep{salin1995corticocortical}.
These many and diverse modulatory influences
from higher onto downstream cortical areas
can inject prior knowledge
at every stage of processing environmental information.
%
Moreover,
both accounts provide a parsimonious explanation why the
human brain's processing load devoted to incoming information decreases
when the environment becomes predictable.
This is because the internal generative
model only requires updates after discrepancies have occurred between
environmental reality and its internally reinstantiated representation.
Increased computation resources are however allocated
when unknown stimuli or
unexpected events are encountered by the organism.
The predictive coding and MDP account hence
naturally evoke a mechanism of brain plasticity in that
neuronal wiring gets increasingly adapted
when faced by unanticipated environmental challenges.

While sensory experience is a constructive process from both views,
the predictive coding account frames
sensory perception of the external world
as a generative experience due to the modulatory top-down influence at
various stages of sensory input processing.
This generative top-down design is replaced in our MDP view of the DMN
by a sequential decision-making framework.
Further,
the hierarchical processing aspect from predictive coding
is re-expressed in our account in form of
nested prediction of probable upcoming actions, states, and outcomes.
While both accounts capture the consequences of action,
the predictive coding account is typically explained without
explicit parameterization of the agent's time horizon and
has a tendency to be presented as emphasizing prediction about the
immediate future.
In the present account, the horizon of that
look into the future is made explicit in the $\gamma$ parameter
of the Bellman equation.
% \item \suggestremove{Mismatch negativity of PC is immanent in reward prediction error
% by the difference between actually predicted reward of an action given
% a state and the reward predicted by non-linear regression with graded
% discounting of future rewards.}
Finally,
the process of adapting the neuronal connections
for improved top-down modulation
takes the concrete form of stochastic gradient computation and
back-propagation in our MDP implementation.
It is however important to note that
the neurobiological plausibility of
the back-propagation procedure is controversal
\citep{goodfellow2016deep}.

@@

\subsection{Semantic hypothesis}
Another frequently proposed cognitive account to explain DMN function revolves
around forming logical associations and abstract analogies between
experiences and conceptual knowledge derived from past behavior
\citep{bar2007proactive, binder1999conceptual, constantinescu2016organizing}.
Analogies might naturally tie incoming new sensory stimuli to
explicit world knowledge (i.e., semantics)
\citep{bar2009proactive}.
The encoding of complex environmental features could thus be facilitated
by association to known similar states.
%
Going beyond isolated meaning and concepts extracted from the world,
semantic building blocks may need to get recombined to enable
mental imagery of non-existing scenarios.
As such, semantic knowledge would be a prerequisite for optimizing behavior
by constantly simulating possible future scenarios
\citep{boyer2008evolutionary, binder2011neurobiology}.
Such cognitive processes can afford
the internal construction and elaboration of necessary information
that is not presented in the surrounding sensory environment
by recombining building blocks of
concept knowledge and episodic memories
\citep{hassabis2009construction}.
Indeed, in aging humans, remembering the past and imaging the future
equally decreased in the level of detail and were associated with
concurrent deficits in forming and integrating relationships between
items \citep{addis2008age, spreng2006temporal}.
Further,
% a constructive account can explain the reciprocal relationship
% between an egocentric first person perspective and
% an allocentric bird’s eye perspective immersed in
% self-reflection, semantic associations, and autobiographical memories.
% %
% Cognitive aspects of egocentric-allocentric switching
episodic memory, language, problem solving,
planning, estimating other others' thoughts, and spatial navigation
represent neural processes that are likely to
build on abstract world knowledge and logical associations
for integrating the constituent elements in rich and coherent mental scenes
\citep{schacter2007remembering}.
Such scene construction processes could contribute to interpreting the
present and foretelling the future.
Further,
mental scene construction has been proposed
to imply a distinction between
engagement in the sensory environment
and internally generated mind-wandering
\citep{buckner2007self}.
These investigators stated that
``A computational model [...] will probably require a form of
regulation by which perception of the current world is suppressed
while simulation of possible alternatives are constructed,
followed by a return to perception of the present.''.
% The semantic hypothesis is for instance supported by evidence in animals that
% could learn a \textit{cognitive map of the environment}
% and exploit it later for other means \citep{tolman1948cognitive}.

@@

In particular,
environmental stimuli especially important for humans are frequently of
social nature. This is probably unsurprising
given that
the complexity of the social systems
is a likely to be a human-defining property
\citep{tomasello2009cultural}.
According to the ``social brain hypothesis'',
the human brain has especially been shaped for
forming and maintaining increasingly complex
social systems,
which allows solving ecological problems by means of social relationships
\citep{whiten1988machiavellian}.
Indeed, social topics amounted to roughly
two thirds of human everyday communication \citep{dunbar1997human},
while
mind-wandering at daytime and dreams during sleep
are rich in stories about people and
the complex relationships between them.
%
In line with this, the DMN was argued to be specialized in
continuous processing of social information as a
physiological baseline of human brain function
\citep{schilbach2008minds}. This view was later challenged by observing
analogues of the DMN in monkeys \citep{mantini2011default},
cats \citep{popa2009contrasting},
and rats \citep{lu2012rat}, three species with
social-cognitive capacities that are supposedly less advanced than in humans.

@@

% CRITICISM
Despite its popularity, criticism against the FEP has arisen over the years,
some of which outlined in the following.
The main algorithm for minimizing free energy $F_G^R(\x)$ is the \textit{wake-sleep algorithm}
\citep{dayan1995helmholtz}. As these authors noted,
a crucial drawback of the wake-sleep algorithm is that it involves a pair
of forward (generation) and backward (recognition) models that
together do not correspond to optimization of (a bound of) the marginal likelihood.
The brain may therefore be unlikely to implement a variant of the wake-sleep algorithm.
The recent theory of
\textit{variational auto-encoders} (VAEs) \citep{kingma2013auto} might provide an efficient alternative to the wake-sleep algorithm. VAEs
overcome a number of the technical limits of the wake-sleep algorithm
by using a reparametrization trick. For instance, unlike the wake-sleep algorithm
for minimizing free energy, VAEs can be efficiently trained via back-propagation of prediction errors.

@@

\section{Conclusion}
Which brain function could be important enough
for the existence and survival of the human species
to justify constantly high energy costs?
MDPs motivate an attractive
formal account how the human association cortex
might implement multi-sensory representation and control of the environment to
optimize the organism's interaction with the world.
% by obtaining reward and avoiding pain.
This idealized process model explains
a number of previous experimental observations in the
DMN by simple but non-trivial mechanisms.
%
From the view of a Markovian sequential decision process,
human behavior unfolds by integrating action outcomes from stored past events
and extrapolation to
upcoming events for guiding action choice in the present context.
MDPs also provide a mathematical formalism how
opportunity in the environment can be recursively exploited
when confronted with challenging decisions.
%
This functional interpretation may be well compatible with the DMN's
poorly understood involvement across
autobiographical memory recall, problem solving,
abstract reasoning, social cognition,
as well as delay discounting and self-related prospection.
Improvement of the internal world representation
by injecting stochasticity into the recall of past
actions and the estimation of action outcomes may explain why
highly accurate memories have been disfavored in human evolution
and why human creativity might be adaptive.

>>>>

\subsubsection{Optimal control via stochastic gradient descent in the DMN}
Efficient learning of the entire set of model parameters can effectively be achieved
via stochastic \textit{gradient descent}, a universal algorithm for finding
local minima based on the first derivative of the optimization objective.
Stochastic here means that the true gradient is estimated from batches of training samples,
which, in our case, corresponds to blocks of experience from the replay memory:

@@

\begin{figure}[!h]
  \includegraphics[width=.9\linewidth]{VennDiagram-2017-09-08.pdf}
  \caption{\textbf{Where does Markov Decision Process fit in the theoretical landscape?}}
  This Venn diagram summarizes the relationship between four existing accounts of DMN: in blue colors, the Free Energy Principle \& Predictive Coding, more related to the ”How” DMN works, and in yellow, the Sentinel \& the Semantic hypotheses, more related to "What" DMN does. Our account of DMN by Markov Decision Process is at the intersection of those accounts and thus corresponds to a "Process model" meaning a model of both "How" and "What" a process occur; here the mental processes and how it relate to observed behavior. For each account, 4 icons describe the core concepts of Time horizon, Neural plasticity, Hierarchy, and Behavioral goal. The "Time horizon" should be understood as a weighting of the expected future outcomes shorter to longer terms into present decision making. Then, "online monitoring" of the environment means detecting and processing salient cues in the sensory environment. The "Predictive Coding" is understood here as the joint minimization of feedforward prediction error and feedback modulation to optimize neural processing; there is thus no notion of an agent with behavioral goal.}
  \label{fig:VennDiagram}
\end{figure}

\section{Relation to existing accounts}

\subsection{Predictive coding hypothesis}
Predictive coding mechanisms
\citep{clark2013whatever, friston2008hierarchical}
are a frequently evoked idea in the context of default mode function
\citep{bar2007}.
Cortical responses are explained as
emerging from continuous functional interaction between
higher and lower levels of the neural processing hierarchy.
Feed-forward sensory processing is constantly calibrated by
top-down modulation from more multi-sensory and associative brain regions
further away from primary sensory cortical regions.
The dynamic interplay between cortical processing levels
may enable learning about aspects of the world by reconciling
gaps between fresh sensory input and expectations computed
based on stored prior information.
At each stage of neural processing,
an internally generated prediction of aspects of environmental sensations is
directly compared against the actual environmental input.
A prediction error at one of the processing levels
incurs plasticity changes of neuronal projections
(i.e., adapting model parameters)
to allow for gradually improved future prediction of the environment.
In this way,
the predictive coding hypothesis offers explanations for
the constructive, non-deterministic nature of sensory perception
\citep{friston2010free, buzsaki2006rhythms} and
the intimate relation of motor action to sensory expectations
\citep{wolpert1995internal, kording2004bayesian}.
Contextual integration of sensorimotor perception-action cycles
may be maintained by top-down modulation
using a-priori information about the environment.

In short, predictive coding processes
conceptualize updates of the internal representation of the environment
to best accommodate and prepare the organism
for processing the constant influx of sensory stimulation and
performing action on the environment.
There are hence a number of common properties
between the predictive coding account
and the proposed formal account of DMN function based on MDPs.
Importantly,
a generative model of how perceived sensory cues arise in the
world would be incorporated into
the current neuronal wiring.
Further,
both functional accounts are plausibilized by
neuroscientific evidence that suggest
the human brain to be a ``statistical organ'' \citep{friston2014phantastic}
with the biological purpose to
generalize from the past to new experiences.
Neuroanatomically, axonal
back projections indeed outnumber by far the axonal input projections
existing in the monkey and probably also human brain
\citep{salin1995corticocortical}.
These many and diverse modulatory influences
from higher onto downstream cortical areas
can inject prior knowledge
at every stage of processing environmental information.
%
Moreover,
both accounts provide a parsimonious explanation for why the
human brain's processing load devoted to incoming information decreases
when the environment becomes predictable.
This is because the internal generative
model only requires updates after discrepancies have occurred between
environmental reality and its internally reinstantiated representation.
Increased computation resources are however allocated
when unknown stimuli or
unexpected events are encountered by the organism.
The predictive coding and MDP account hence
naturally evoke a mechanism of brain plasticity in that
neuronal wiring gets increasingly adapted
when faced by unanticipated environmental challenges.

While sensory experience is a constructive process from both views,
the predictive coding account frames
sensory perception of the external world
as a generative experience due to the modulatory top-down influence at
various stages of sensory input processing.
This generative top-down design is replaced in our MDP view of the DMN
by a sequential decision-making framework.
Further,
the hierarchical processing aspect from predictive coding
is re-expressed in our account in the form of
nested prediction of probable upcoming actions, states, and outcomes.
While both accounts capture the consequences of action,
the predictive coding account is typically explained without
explicit parameterization of the agent's time horizon and
has a tendency to be presented as emphasizing prediction about the
immediate future.
In the present account, the horizon of that
look into the future is made explicit in the $\gamma$ parameter
of the Bellman equation.
% \item \suggestremove{Mismatch negativity of PC is immanent in reward prediction error
% by the difference between actually predicted reward of an action given
% a state and the reward predicted by non-linear regression with graded
% discounting of future rewards.}
Finally,
the process of adapting the neuronal connections
for improved top-down modulation
takes the concrete form of stochastic gradient computation and
back-propagation in our MDP implementation.
It is however important to note that
the neurobiological plausibility of
the back-propagation procedure is controversial
\citep{goodfellow2016deep}.

@@

\subsection{Semantic hypothesis}
Another frequently proposed cognitive account to explain DMN function revolves
around forming logical associations and abstract analogies between
experiences and conceptual knowledge derived from past behavior
\citep{bar2007proactive, binder1999conceptual, constantinescu2016organizing}.
Analogies might naturally tie incoming new sensory stimuli to
explicit world knowledge (i.e., semantics)
\citep{bar2009proactive}.
The encoding of complex environmental features could thus be facilitated
by association to known similar states.
%
Going beyond isolated meaning and concepts extracted from the world,
semantic building blocks may need to get recombined to enable
mental imagery of non-existing scenarios.
As such, semantic knowledge would be a prerequisite for optimizing behavior
by constantly simulating possible future scenarios
\citep{boyer2008evolutionary, binder2011neurobiology}.
Such cognitive processes can afford
the internal construction and elaboration of necessary information
that is not presented in the surrounding sensory environment
by recombining building blocks of
concept knowledge and episodic memories
\citep{hassabis2009construction}.
Indeed, in aging humans, remembering the past and imagining the future
equally decreased in the level of detail and were associated with
concurrent deficits in forming and integrating relationships between
items \citep{addis2008age, spreng2006temporal}.
Further,
% a constructive account can explain the reciprocal relationship
% between an egocentric first person perspective and
% an allocentric bird’s eye perspective immersed in
% self-reflection, semantic associations, and autobiographical memories.
% %
% Cognitive aspects of egocentric-allocentric switching
episodic memory, language, problem solving,
planning, estimating others' thoughts, and spatial navigation
represent neural processes that are likely to
build on abstract world knowledge and logical associations
for integrating the constituent elements in rich and coherent mental scenes
\citep{schacter2007remembering}.
Such scene construction processes could contribute to interpreting the
present and foretelling the future.
Further,
mental scene construction has been proposed
to imply a distinction between
engagement in the sensory environment
and internally generated mind-wandering
\citep{buckner2007self}.
These investigators stated that
``A computational model [...] will probably require a form of
regulation by which perception of the current world is suppressed
while simulation of possible alternatives are constructed,
followed by a return to perception of the present.''.
% The semantic hypothesis is for instance supported by evidence in animals that
% could learn a \textit{cognitive map of the environment}
% and exploit it later for other means \citep{tolman1948cognitive}.

@@

In particular,
environmental stimuli especially important for humans are frequently of
social nature. This is probably unsurprising
given that
the complexity of the social systems
is likely to be a human-defining property
\citep{tomasello2009cultural}.
According to the ``social brain hypothesis'',
the human brain has especially been shaped for
forming and maintaining increasingly complex
social systems,
which allows solving ecological problems by means of social relationships
\citep{whiten1988machiavellian}.
Indeed, social topics amounted to roughly
two thirds of human everyday communication \citep{dunbar1997human},
while
mind-wandering at daytime and dreams during sleep
are rich in stories about people and
the complex relationships between them.
%
In line with this, the DMN was argued to be specialized in
continuous processing of social information as a
physiological baseline of human brain function
\citep{schilbach2008minds}. This view was later challenged by observing
analogues of the DMN in monkeys \citep{mantini2011default},
cats \citep{popa2009contrasting},
and rats \citep{lu2012rat}, three species with
social-cognitive capacities that are supposedly less advanced than in humans.

@@

% CRITICISM
Despite its popularity, criticism against the FEP has arisen over the years,
some of which is outlined in the following.
The main algorithm for minimizing free energy $F_G^R(\x)$ is the \textit{wake-sleep algorithm}
\citep{dayan1995helmholtz}. As these authors noted,
a crucial drawback of the wake-sleep algorithm is that it involves a pair
of forward (generation) and backward (recognition) models that
together do not correspond to optimization of (a bound of) the marginal likelihood.
The brain may therefore be unlikely to implement a variant of the wake-sleep algorithm.
The recent theory of
\textit{variational auto-encoders} (VAEs) \citep{kingma2013auto} might provide an efficient alternative to the wake-sleep algorithm. VAEs
overcome a number of the technical limits of the wake-sleep algorithm
by using a reparametrization maneuver. For instance, unlike the wake-sleep algorithm
for minimizing free energy, VAEs can be efficiently trained via back-propagation of prediction errors.

@@

\section{Conclusion}
Which brain function could be important enough
for the existence and survival of the human species
to justify constantly high energy costs?
MDPs motivate an attractive
formal account of how the human association cortex
might implement multi-sensory representation and control of the environment to
optimize the organism's interaction with the world.
% by obtaining reward and avoiding pain.
This idealized process model explains
a number of previous experimental observations in the
DMN by simple but non-trivial mechanisms.
%
From the view of a Markovian sequential decision process,
human behavior unfolds by integrating action outcomes from stored past events
and extrapolation to
upcoming events for guiding action choice in the present context.
MDPs also provide a mathematical formalism how
opportunity in the environment can be recursively exploited
when confronted with challenging decisions.
%
This functional interpretation may well be compatible with the DMN's
poorly understood involvement across
autobiographical memory recall, problem solving,
abstract reasoning, social cognition,
as well as delay discounting and self-related prospection.
Improvement of the internal world representation
by injecting stochasticity into the recall of past
actions and the estimation of action outcomes may explain why
highly accurate memories have been disfavored in human evolution
and why human creativity might be adaptive.
